\documentclass[11pt,fleqn]{article}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in]{geometry}
%use helvetica 11pt
%\usepackage{helvet}
%\renewcommand{\familydefault}{\sfdefault} 
\usepackage{xspace}
%\usepackage{cite}
%\usepackage[labelfont={footnotesize,bf},font={footnotesize}]{caption}

\pagestyle{empty}

%\graphicspath{{./Figures/}}
%\DeclareGraphicsExtensions{.pdf}
\begin{document}
\section*{Project Facilities}
\subsection*{Office Space}
Both Investigators have their own offices which are less than a 10 minute walk from each other.
PI's has adequate office space for Post Doctoral, Graduate Student, and Undergraduate Student researchers in computation lab next door to his office.
All offices are air-conditioned and have internet connections available.

\subsection*{Computer Resources}
Head PI has three multi-core computer workstations (8, 8, and 12 cores) available for use by personnel on this project.
For more computationally intensive work, students and faculty have secondary priority to a +1200 core cluster, Newton, as part of  joint effort between OIT and other departments at the University.
We have requested funds for an additional 64 cores that will be added to Newton and for which we will have first priority. 
In addition, the Head-PI is a senior personnel at the recently established National Institute for Mathematical and Biological Synthesis (NIMBioS) which is hosted by the University of Tennessee.
NIMBioS runs a 128 core cluster as well as an X-Grid cluster consisting of 40 networked Mac workstations.
In addition the University is in charge of maintaining a 18,000 core Cray XT4 supercomputer as part of the National Institute for Computational Sciences (NICS).
Members of the University have priority for 10\% of this cluster's computation time.



\end{document}
